---
title: 'Project #2 - R Data Analysis'
output: html_document
---


```{r}
library(dplyr) 
library(tidyr) 
library(ggplot2)
library(tidyverse)

source('func.R')
```


─ Conflicts ────────────────────────────────────────────────────────────────────────────────────────────────────
───── tidyverse_conflicts() ──
✖ dplyr::between()   masks data.table::between()
✖ dplyr::filter()    masks stats::filter()
✖ dplyr::first()     masks data.table::first()
✖ dplyr::lag()       masks stats::lag()
✖ dplyr::last()      masks data.table::last()
✖ purrr::transpose() masks data.table::transpose()


## Data source
Data source:

*** I. mobility data
- mobility: https://www.google.com/covid19/mobility/


*** II. county-level COVID data
- CDC: https://data.cdc.gov/Public-Health-Surveillance/United-States-COVID-19-County-Level-Data-Sources/7pvw-pdbr
- NYC cases: https://www1.nyc.gov/site/doh/covid/covid-19-data.page
- NYC vaccine: https://www1.nyc.gov/site/doh/covid/covid-19-data-vaccines.page#doses


*** III. county-level census data
- 2020 CENSUS RESULTS FOR NEW YORK CITY.pdf
- nyc.gov/planning August 2021


## Data pre-processing

### I. Mobility data
#### 1. Load original data
```{r}
### load original data
us_2020 <- read.csv("./data/Region_Mobility_Report_CSVs/2020_US_Region_Mobility_Report.csv", 
                    stringsAsFactors = FALSE,na="")

us_2021 <- read.csv("./data/Region_Mobility_Report_CSVs/2021_US_Region_Mobility_Report.csv", 
                    stringsAsFactors = FALSE,na="")

us_2022 <- read.csv("./data/Region_Mobility_Report_CSVs/2022_US_Region_Mobility_Report.csv", 
                    stringsAsFactors = FALSE,na="")

```


```{r}
### crosscheck dimensions and column names. Make sure 1) no. of columns and 2) column names are matched
dim(us_2020)
dim(us_2021)
dim(us_2022)

colnames(us_2020) == colnames(us_2021)
colnames(us_2021) == colnames(us_2022)

```

```{r}
### check column names
colnames(us_2020)
```


```{r}
### preliminary look of the data
#View(us_2020)
head(us_2020)
```


```{r}
### check type and range of each column - 1
str(us_2020)
str(us_2021)
str(us_2022)
```
```{r}
### check type and range of each column - 2
summary(us_2020)
#summary(us_2021)
#summary(us_2022)
```


```{r}
### understand columns
#unique(us_2020$sub_region_1) # state
#unique(us_2020$sub_region_2) # county
#unique(us_2020$metro_area) # all NA
```

#### 2. subset dataframe for anallysis use
```{r}
### subset NYC data

nyc_county = c("New York County", # Manhattan
               "Kings County", # Brooklyn
               "Bronx County", # Bronx
               "Richmond County", # Staten Island
               "Queens County") # Queens

nyc_2020 <- us_2020 %>% 
     filter(  (sub_region_1 == "New York")  & (sub_region_2 %in% nyc_county))

nyc_2021 <- us_2021 %>% 
     filter(  (sub_region_1 == "New York")  & (sub_region_2 %in% nyc_county))

nyc_2022 <- us_2022 %>% 
     filter(  (sub_region_1 == "New York")  & (sub_region_2 %in% nyc_county))

dim(nyc_2020) # 1605   15
dim(nyc_2021) # 1825   15
dim(nyc_2022) # 980  15

```

```{r}
### check dim 
nyc_2020 %>% group_by(sub_region_2) %>% summarise(n())
nyc_2021 %>% group_by(sub_region_2) %>% summarise(n())
nyc_2022 %>% group_by(sub_region_2) %>% summarise(n())

### check NAs in date,sub_region_2, 6 categories
colSums(is.na(nyc_2020)) # 8 - parks_percent_change_from_baseline 
colSums(is.na(nyc_2021)) # no NA
colSums(is.na(nyc_2022)) # no NA

nyc_2020 %>%
  filter(is.na(parks_percent_change_from_baseline))  # Richmond County's park mobility data has 8 NAs (2020-8-18 ~ 2020-9-7)
```


```{r}
### combine annual mobility data
mobility_nyc <- rbind(nyc_2020, nyc_2021, nyc_2022)

#str(mobility_nyc$date)

```

```{r}
### check dim and each column
dim(mobility_nyc) # 4410   15
summary(mobility_nyc)
str(mobility_nyc)
```


```{r}
### check numbers of data
mobility_nyc %>% group_by(sub_region_2) %>% summarise(n())
```

NYC County (Borough)

New York County (Manhattan)
Kings County (Brooklyn)
Bronx County (The Bronx)
Richmond County (Staten Island)
Queens County (Queens)


```{r}
county_nyc = c("New York County",
           "Kings County",
           "Bronx County",
           "Richmond County", 
           "Queens County")

boro_nyc = c("Manhattan",
             "Brooklyn",
             "Bronx",
             "Staten Island",
             "Queens")
```


Change date type and subset
```{r}
### change data type & subset by date ()
mobility_nyc <- mobility_nyc %>%
  mutate(date = as.Date(date)) %>%
  filter(date < "2022-7-1")
```


#### 3. pre-processing
Add feature - weekday
```{r}
### add feature: weekday
#install.packages("lubridate")
library(lubridate)
mobility_nyc$weekday <- wday(mobility_nyc$date,week_start=1) #,label=TRUE)

#mobility_nyc$week <- wday(mobility_nyc$date,week_start=1,label=TRUE)
#mobility_nyc$week <- wday(mobility_nyc$date,label=TRUE)

```


Change column names
```{r}
mobility_nyc = mobility_nyc %>% 
  rename(state = sub_region_1,
         borough = sub_region_2,
         retail_and_recreation = retail_and_recreation_percent_change_from_baseline,
         grocery_and_pharmacy = grocery_and_pharmacy_percent_change_from_baseline,
         parks = parks_percent_change_from_baseline,
         transit_stations = transit_stations_percent_change_from_baseline,
         workplaces = workplaces_percent_change_from_baseline,
         residential = residential_percent_change_from_baseline)


```



```{r}
### split workplaces and residentials by weekdays

mobility_nyc  <- mobility_nyc  %>%
  mutate(workplaces_weekdays = case_when(weekday >=1 & weekday <=5   ~ workplaces),
         
         workplaces_weekends = case_when(weekday == 6 | weekday == 7  ~ workplaces),
         
         residential_weekdays = case_when(weekday >=1 & weekday <=5   ~ residential),
         
         residential_weekends = case_when(weekday == 6 | weekday == 7  ~ residential))


```


Replace county name by borough names
```{r}
mobility_nyc <- mobility_nyc %>%
  mutate( borough = case_when( 
    borough == "Bronx County" ~ "Bronx",
    borough == "Kings County" ~ "Brooklyn",
    borough == "New York County" ~ "Manhattan",
    borough == "Queens County" ~ "Queens",
    borough == "Richmond County" ~ "Staten Island"))

```


Save a copy
```{r}
### save a copy of pre-processed mobility data
#write.csv(mobility_nyc, "./data/mobility_nyc.csv",row.names=FALSE)

#df_mobility_nyc <- mobility_nyc
#mobility_nyc <- df_mobility_nyc
#dim(df_mobility_nyc) # 4335 15

```



### II. COVID data
#### 1. Load original data
```{r}
### load original data
cases_nyc <- read.csv("./data/covid/nyc_data_by_day.csv", 
                     stringsAsFactors = FALSE, na="")

vx_nyc <- read.csv("./data/covid/nyc_trends_byboro.csv", 
                     stringsAsFactors = FALSE, na="")

census_nyc <- read.csv("./data/nyc_2020_census.csv", 
                     stringsAsFactors = FALSE, na="")
```

#### Case/hosp/death data
```{r}
### check dimension
dim(cases_nyc) # 868  67
dim(vx_nyc) # 582  37
dim(census_nyc) # 5 3
```
```{r}
### check type and range of each column
str(cases_nyc)
summary(cases_nyc)

str(vx_nyc) # [To-do] some mobility data are chr type, including "<10"
summary(vx_nyc)

str(census_nyc)
summary(census_nyc)
```


```{r}
### check NAs
colSums(is.na(cases_nyc)) # no NAs
colSums(is.na(vx_nyc)) # no NAs
colSums(is.na(census_nyc)) # no NAs
```

#### 2. subset (by column)
```{r}
### subset for analysis (by columns)

#colnames(cases_nyc)

case_col = colnames(cases_nyc)[c(1,
         12,14,15,
         23,25,26,
         34,36,37,
         45,47,48,
         56,58,59)]

cases_nyc_v1 <- cases_nyc %>% 
  select(all_of(case_col))

colnames(cases_nyc_v1)
```
#### 3. data processing
```{r}
### data processing
#### 1. change data type - date
#### 2. modify column names
#### 3. remove NAs - no NAs
#### 4. subset by date for analysis period (2020.2.15-2022.6.30)

cases_nyc_v2 <- cases_nyc_v1 %>%
  mutate(date_of_interest = as.Date(date_of_interest, format="%m/%d/%Y")) %>%
  rename(date = date_of_interest) %>%
  pivot_longer(cols = -date) %>%
  separate(col = name, into = c("borough","ctype") , sep =2 ) %>% 
  pivot_wider(id_cols = c(date,borough), names_from = ctype, values_from = value) %>%
  mutate( borough = case_when( borough == 'BK' ~ 'Brooklyn',
                              borough == 'BX' ~ 'Bronx',
                              borough == 'MN' ~ 'Manhattan',
                              borough == 'QN'~ 'Queens',
                              borough == 'SI' ~ 'Staten Island')) %>%
  rename(case_count = '_CASE_COUNT' ,
         hosp_count = '_HOSPITALIZED_COUNT' ,
         death_count = '_DEATH_COUNT') %>%
  rename_with(tolower)%>%
  filter(date <"2022-7-1")


#View(cases_nyc_v2)
str(cases_nyc_v2)
summary(cases_nyc_v2)
cases_nyc_v2 %>% group_by(borough)  %>% summarise(n())
```



#### vaccine
```{r}
### check dimension
dim(vx_nyc) # 582  37
```

```{r}
### check type and range of each column
str(vx_nyc) # [To-do] some mobility data are chr type, including "<10"
summary(vx_nyc)
```


```{r}
### check NAs
colSums(is.na(vx_nyc)) # no NAs
```


#### 2. subset by column
#### 3. data processing
```{r}
colnames(vx_nyc)
str(colnames(vx_nyc))

### subset (by columns), change data type, remove NAs
vx_nyc_v1 <- vx_nyc %>% 
  select(-2,-3,-4,-5,-6,-7) %>%
  mutate(DATE = as.Date(DATE)) %>%
  mutate_if(is.character,   funs(ifelse(.=="<10", "0", .))) %>%
  mutate_if(is.character,  as.integer)

colnames(vx_nyc_v1)
dim(vx_nyc_v1)
str(vx_nyc_v1)
```


```{r}
### data preperation:
#### 1. change data type (previous)
#### 2. modify column names
#### 3. remove NAs (previous)
#### 4. subset by date for analysis period (2020.2.15-2022.6.30)


vx_nyc_v2 <- vx_nyc_v1 %>%
  pivot_longer(cols = -DATE) %>%
  separate(col = name, into = c("borough","vtype") , sep =2 ) %>% 
  pivot_wider(id_cols = c(DATE,borough), names_from = vtype, values_from = value) %>%
  mutate( borough = case_when( borough == 'BK' ~ 'Brooklyn',
                              borough == 'BX' ~ 'Bronx',
                              borough == 'MH' ~ 'Manhattan',
                              borough == 'QS'~ 'Queens',
                              borough == 'SI' ~ 'Staten Island')) %>%
  rename(vx_fully = '_COUNT_FULLY_CUMULATIVE' ,
          vx_plus1 = '_COUNT_1PLUS_CUMULATIVE' ,
          vx_additional = '_COUNT_ADDITIONAL_CUMULATIVE',
          vx_fully_perc = '_PERC_FULLY' ,
          vx_plus1_perc = '_PERC_1PLUS' ,
          vx_additional_perc = '_PERC_ADDITIONAL' ) %>%
  rename_with(tolower) %>%
  filter(date <"2022-7-1")

str(vx_nyc_v2)
summary(vx_nyc_v2)
vx_nyc_v2 %>% group_by(borough)  %>% summarise(n()) #564
```

#### 4. case & vx data combined
```{r}
dim(cases_nyc_v2) #4265    5
dim(vx_nyc_v2) #2820    8

covid_nyc_1 = left_join(cases_nyc_v2, vx_nyc_v2, by = c("date","borough"))
covid_nyc = left_join(covid_nyc_1, census_nyc, by = "borough")

#View(covid_nyc)
summary(covid_nyc)
dim(covid_nyc) #4265 13

```

#### 5. Generate features
```{r}
#### generate features
#colnames(covid_nyc)

# 1 - percent count
covid_nyc <- covid_nyc %>%
  mutate(case_count_perc = (case_count/population)*100,
         hosp_count_perc = (hosp_count/population)*100,
         death_count_perc = (death_count/population)*100)

colnames(covid_nyc)
```

Save a copy of pre-processed COVID data
```{r}
#write.csv(cases_nyc_v2, "./data/covid_nyc.csv",row.names=FALSE)
#df_cases_nyc <- cases_nyc_v2
#cases_nyc_v2 <- df_cases_nyc

#write.csv(vx_nyc_v2, "./data/covid_nyc.csv",row.names=FALSE)
#df_vx_nyc <- vx_nyc_v2
#vx_nyc_v2 <- df_vx_nyc

write.csv(covid_nyc, "./data/covid_nyc.csv",row.names=FALSE)
#df_covid_nyc <- covid_nyc
#covid_nyc <- df_covid_nyc
```


## Data analysis:

### Part I - By category

#### 1. overall
```{r, fig.width = 20, fig.height = 5, warning=FALSE}
#categories_ = colnames(mobility_nyc)[10:15]
categories_ = colnames(mobility_nyc)[c(10:13,17:20)]

mobility_nyc %>% 
  #group_by(borough) %>%
  #summarise(median = median(retail_and_recreation,na.rm=TRUE),
  #          mean = mean(retail_and_recreation,na.rm=TRUE))%>%
  select(borough,all_of(categories_))%>%
  pivot_longer(cols = -borough)%>%
  ggplot(aes(x=name, y=value, fill=borough),na.rm=TRUE) +
  geom_violin(position='dodge') + 
  labs(title = "Distribution of Mobility Change from Baseline (%) by Category", x="Date",y="Percentage Change (%)")+
  guides(fill=guide_legend("Borough"))

```



```{r, fig.width = 10, fig.height = 5}
library(ggpubr)
source('func.R')

library(patchwork)

```


"Retail and Recreation"
"Grocery and Pharmacy"
"Parks"
"Transit Stations"
"Workplaces"
"Residential"

*1
```{r, fig.width = 10, fig.height = 5}
plots_by_category("Retail and Recreation")
```

* 2
```{r, fig.width = 10, fig.height = 5}
plots_by_category("Grocery and Pharmacy")
```

* 3
```{r, fig.width = 10, fig.height = 5}
plots_by_category("Parks")
```

* 4
```{r, fig.width = 10, fig.height = 5}
plots_by_category("Transit Stations")
```

* 5
```{r, fig.width = 10, fig.height = 5}
plots_by_category("Workplaces")
```
* 5.weekdays
```{r, fig.width = 10, fig.height = 5}
plots_by_category("Workplaces Weekdays")
```


* 5.weekends
```{r, fig.width = 10, fig.height = 5}
plots_by_category("Workplaces Weekends")
```



* 6
```{r, fig.width = 10, fig.height = 5}
plots_by_category("Residential")
```


* 6.weekdays
```{r, fig.width = 10, fig.height = 5}
plots_by_category("Residential Weekdays")
```

* 6.weekends
```{r, fig.width = 10, fig.height = 5}
plots_by_category("Residential Weekends")
```




#### NOT IN USE - sidenote
```{r}
mobility_nyc %>%
  filter(borough=="Manhattan")  %>%    
  ggplot() +
    geom_point(aes(x=date,y=workplaces), size=1, alpha=0.4) +
    labs(title = "Workplace Change from Baseline(%)", x="Date",y="Percentage Change (%)") +
    #ylim(-100,100) +
    geom_hline(yintercept=0, linetype="dashed", color = "red") #+#, size=1) +
    #geom_smooth(method = lm)
```



```{r}
mobility_nyc %>%
  filter(borough=="Manhattan")  %>%    
  ggplot() +
    geom_point(aes(x=date,y=parks), size=1, alpha=0.4) +
    labs(title = "Parks Change from Baseline(%)", x="Date",y="Percentage Change (%)") +
    #ylim(-100,100) +
    geom_hline(yintercept=0, linetype="dashed", color = "red") #+#, size=1) +
    #geom_smooth(method = lm)
```

1. Manhattan is the most impacted borough across all categories.
2. Retail & Recreation - shows clear recovery since 2021 but still under pre-covid level. Manhattan is heavily impacted and the other 4 boroughs are similar.
3. Grocery and Pharmacy - least impacted category but still slightly under water.
4. Parks - Seasonality with peaks in summer and early fall time. Brooklyn and Queens have very high variations by seasonality, but mobility is still relatively lower than the baseline period in each year.
5. Transit Stations - Still suffering large decreases from baseline across boroughs.
6. Workplaces - largest impacted category.
7. Residential - increase since COVID. The largest increase was in H1 2020 and shows decreasing trends.


### Part II - Mobility Relationship with Covid (Manhattan)


#### 2.1 Mobility vs COVID stats
```{r}
dim(mobility_nyc) 
dim(covid_nyc)
nyc <- left_join(mobility_nyc,covid_nyc,by=c("date","borough"))

dim(nyc)
```

###### single category of a borough - side analysis
```{r, warning=FALSE}

nyc_1 = nyc %>%
  filter(borough =="Manhattan") %>%
  select(date,case_count,parks) %>%
  pivot_longer(col=-date) %>%
  mutate(
    case_count = ifelse(name=="case_count", value, NA),
    parks =  ifelse(name=="parks", value, NA))

nyc_1$name <- factor(nyc_1$name, levels = c("case_count", "parks"),
                     labels = c("Case Count", "Parks change(%)"))

# df$supp <- factor(df$supp, levels = c("OJ", "VC"),
#                   labels = c("Orange Juice", "Vitamin C")
#                   )

ggplot(nyc_1) + 
  geom_line(aes(x=date, y=case_count),color="red") +
#  ggplot_add(geom_area(data= nyc,aes(x=date, y=vx_fully),fill="red"))+
  geom_line(aes(x=date,y=parks),color="blue") +
  facet_grid(name ~ ., scales="free_y") +
  labs(title = "Manhattan", x="Date",)+
  ylab("")

```

###### all category of a borough
```{r, fig.width = 10, fig.height = 20}
#categories_
categories_colname = colnames(mobility_nyc)[10:15]#[c(10:13,17:20)]
categories_colname

label1 = c()
for (a in categories_colname[1:length(categories_colname)]) {
  label1 = append(label1,
               gsub("And", "&", paste(paste(str_to_title(strsplit(a, "_")[[1]]),collapse = " "),"Change (%)",sep=" ")))
  
}

label1 = gsub("Weekdays", "(Weekdays)", label1)
label1 = gsub("Weekends", "(Weekends)", label1)

label1

```


```{r, fig.width = 10, fig.height = 15}
source('func.R')

#categories_colname = colnames(mobility_nyc)[10:15]
#categories_colname = colnames(mobility_nyc)[10:15]#[c(10:13,17:20)]

boro_plot("Manhattan","Cases")
boro_plot("Manhattan",covid_stat = "Vaccinations")
boro_plot("Queens",covid_stat = "Deaths")

```



#### 2.2 Correlation
```{r}
library(corrplot)
cutoff_date = "2021-11-30"
#df_nyc = left_join(mobility_nyc, covid_nyc, by = c("date","borough"))

dim(mobility_nyc) #[1] 4335   15
dim(covid_nyc) #[1] 4265   11
#dim(df_nyc) #[1] 4335   24

```


```{r, fig.width = 10, fig.height = 8}
source('func.R')

corr_matrix_plot(region="NYC", "2021-11-30")
corr_matrix_plot(region="Manhattan", "2021-11-30")
#corr_matrix_plot(region="Brooklyn", "2021-11-30")

```
1. Residential mobility change is positively correlated with COVID stats, and all other categories are negatively correlated with them.
2. Correlations between mobility change and COVID stats gradually decrease.

### Additional graphs for presentation
```{r}
#{r, fig.width = 4, fig.height = 3}

census_nyc %>% 
  ggplot(aes(x=reorder(borough, -population), y=population, fill=borough)) +
  geom_bar(stat='identity', width = 0.5) +
  #geom_bar(position='dodge', stat='identity')
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"),
        axis.title.y=element_blank(),legend.position="none")+
  guides( y = "none")+  
  scale_y_continuous(labels = scales::comma) +
  geom_text(aes(label = format(population, big.mark = ",")), 
              position = position_stack(vjust = 1.05))+
  geom_text(aes(label = scales::percent(percent/100, big.mark = ",")), 
              position = position_stack(vjust = 0.5),
            color="white")+
  labs(title = "Population, New York City 2020", x="Borough",y="Population")+
  guides(fill=guide_legend("Borough"))+
  theme(plot.title = element_text(size = 18, face = "bold"))


```

```{r, fig.width = 10, fig.height = 6}
nyc %>% 
  group_by(date) %>% 
  summarise(total = sum(case_count, na.rm = TRUE)) %>%
  ggplot(aes(x=date, y=total)) +
  geom_line(size=1,color="blue") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_date(date_breaks = "2 months")+
  scale_y_continuous(labels = scales::comma) +
  labs(title = "New York City Confirmed Cases", x="Date",y="Count")+
  theme(plot.title = element_text(size = 18, face = "bold"))

  
```
